---
title: "jmatrixsc"  
author: "Juan Domingo"  
output: rmarkdown::html_vignette  
bibliography: scellpam.bib
vignette: >
  %\VignetteIndexEntry{jmatrixsc}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Please, read this  

This is a copy of the vignette of package `jmatrix`. It is included 
here since `jmatrix` is underlying this package and you will need to know 
how to prepare your data to be processed by `scellpam`. But you must 
NOT load package `jmatrix`; all the functions detailed below are 
already included into `scellpam` and are available just by loading it with
`library(scellpam)`.

```{r setup}
library(scellpam)
```

# Purpose  

The package `jmatrix` (@R-jmatrix) was originally conceived as a tool for other
packages, namely `parallelpam` (@R-parallelpam) and `scellpam` (@R-scellpam) which
needed to deal with very big matrices which might not fit in the memory of the
computer, particularly if their elements are of type `double` (in most modern
machines, 8 bytes per element) whereas they could fit if they were matrices of
other  data types, in particular of floats (4 bytes per element).

Unfortunately, R is not a strongly typed language. Double is the default 
type in R and it is not easy to work with other data types. Trials like 
the package float (@R-float) have been done, but to use them you have to coerce a
matrix already loaded in R memory to a float matrix, and then you can delete it.
But, what happens if you computer has not memory enough to hold the matrix in
the first place?. This is the problem this package tries to address.

Our idea is to use the disk as temporarily storage of the matrix in a
file with a internal binary format (`jmatrix` format). This format has a 
header of 128 bytes with information like type of matrix (full, sparse or
symmetric), data type of each element (char, short, int, long, float, double or
long double), number of rows and columns and endianness; then comes the content 
as binary data (in sparse matrices zeros are not stored; in symmetric matrices
only the lower-diagonal is stored) and finally the metadata (currently, 
names for rows/columns if needed and an optional comment).

Such files are created and loaded by functions written in C++ which are 
accessible from `R` with Rcpp (@R-Rcpp). The file, once loaded, uses strictly the
needed memory for its data type and can be processed by other C++ functions
(like the PAM algorithm or any other numeric library written in C++) also
from inside `R`.

The matrix contained in a binary data file in `jmatrix` format cannot be 
loaded directly in `R` memory as a `R` matrix (that would be impossible, 
anyway, since precisely this package is done for the cases in which such 
matrix would NOT fit into the available RAM). Nevertheless, limited access
through some functions is provided to read one or more rows or one or 
more columns as `R` vectors or matrices (obviously, coerced to double).

The package `jmatrix` must not be considered as a final, finished
software. Currently is mostly an instrumental solution to address our needs and
we make available as a separate package just in case it could be useful for
anyone else.

# Workflow

## Debug messages

First of all, the package can show quite informative (but sometimes 
verbose) messages in the console. To turn on/off such messages you can use.
```{r}
ScellpamSetDebug(TRUE,TRUE,TRUE)
# Initially, state of debug is FALSE.
```

## Data storage

As stated before, the binary matrix files should normally be created 
from C++ getting the data from an external source
like a data file in a format used in bioinformatics or a .csv file. 
These files should be read by chunks. As an example,
look at function `CsvToJMat` in package `scellpam`.

As a convenience and only for testing purposes (to be used in this vignette), 
we provide the function `JWriteBin` to write a
R matrix as a `jmatrix` file.
```{r}
# Create a 6x8 matrix of random values
Rf <- matrix(runif(48),nrow=6)
# Set row and column names for it
rownames(Rf) <- c("A","B","C","D","E","F")
colnames(Rf) <- c("a","b","c","d","e","f","g","h")
# Let's see the matrix
Rf
# and write it as the binary file Rfullfloat.bin
JWriteBin(Rf,"Rfullfloat.bin",dtype="float",dmtype="full",
          comment="Full matrix of floats")
# Also, you can write it with double data type:
JWriteBin(Rf,"Rfulldouble.bin",dtype="double",dmtype="full",
          comment="Full matrix of doubles")
```

To get information about the stored file the function `JMatInfo` is provided.
Of course, this funcion does not read the
complete file in memory but just the header.
```{r}
# Information about the float binary file
JMatInfo("Rfullfloat.bin")
# Same information about the double binary file
JMatInfo("Rfulldouble.bin")
```

A jmatrix binary file can be exported to .csv/.tsv table. This is done with
the function `JMatToCsv`
```{r}
# Create a 6x8 matrix of random values
Rf <- matrix(runif(48),nrow=6)
# Set row and column names for it
rownames(Rf) <- c("A","B","C","D","E","F")
colnames(Rf) <- c("a","b","c","d","e","f","g","h")
# Store it as the binary file Rfullfloat.bin
JWriteBin(Rf,"Rfullfloat.bin",dtype="float",dmtype="full",
          comment="Full matrix of floats")
# Save the content of this .bin as a .csv file
JMatToCsv("Rfullfloat.bin","Rfullfloat.csv",csep=",",withquotes=FALSE)
```
The generated file will not have quotes neither around the column names (in its first line)
nor around each row name (at the beginning of each line) since withquotes is FALSE but it
can be set to TRUE for the opposite behavior. Also, a .tsv (tabulator separated values)
would have been generated using csep="\\t".

Also, a jmatrix binary file can also be generated from a .csv/.tsv file.
Such file must have a first line with the names of the columns (possibly surrounded by double quotes,
including a first empty double-quote, since the column of row names has no name itself).
The rest of its lines must start with a string (possibly surrounded by double quotes) with the row
name and the values. In all cases (first line and data lines) each column must be separated
from the next by a separation character (usually, a comma). No separation character must be
added at the end of each line. This format is compatible with the .csv generated by R with
the function `write.csv`.

The function to read .csv files is `CsvToJMat`
```{r}
# Create a 6x8 matrix of random values
Rf <- matrix(runif(48),nrow=6)
# Set row and column names for it
rownames(Rf) <- c("A","B","C","D","E","F")
colnames(Rf) <- c("a","b","c","d","e","f","g","h")
# Save it as a .csv file with the standard R function...
write.csv(Rf,"rf.csv")
# ...and read it to create a jmatrix binary file
CsvToJMat("rf.csv","rf.bin",mtype="full",csep=",",ctype="raw",valuetype="float",transpose=FALSE,comment="Test matrix generated reading a .csv file")
# Let's see the characteristics of the binary file
JMatInfo("rf.bin")
```

### Special note for symmetric matrices:

The parameter mtype="symmetric" will consider the content of the .csv file as a symmetric matrix. This implies that
it must be a square matrix (same number of rows and columns) but the upper-diagonal matrix that must be present
(it does not matter with which values) will be read, and immediately ignored, i.e.: only the lower-diagonal matrix
(including the main diagonal) will be stored.

## Data load  

As stated before, no function is provided to read the whole matrix in 
memory which would contradict the philosophy of this package,
but you can get rows or columns from a file.
```{r}
# Reads row 1 into vector vf. Float values inside the file are
# promoted to double.
(vf<-GetJRow("Rfullfloat.bin",1))
```

Obviously, storage in float provokes a loosing of precision. We have 
observed this not to be relevant for `PAM` (partitioning around medoids)
algorihm but it can be important in other cases. It is the price to pay
for halving the needed space.
```{r}
# Checks the precision lost
max(abs(Rf[1,]-vf))
```

Nevertheless, storing as double obviously keeps the data intact.
```{r}
vd<-GetJRow("Rfulldouble.bin",1)
max(abs(Rf[1,]-vd))
```

Now, let us see examples of some functions to read rows or columns by 
number or by name, or to read several rows/columns as a R matrix.
In all examples numbers for rows and columns are in R-convention (i.e. 
starting at 1)
```{r}
# Read column number 3
(vf<-GetJCol("Rfullfloat.bin",3))
# Test precision
max(abs(Rf[,3]-vf))
# Read row with name C
(vf<-GetJRowByName("Rfullfloat.bin","C"))
# Read column with name c
(vf<-GetJColByName("Rfullfloat.bin","c"))
# Get the names of all rows or columns as vectors of R strings
(rn<-GetJRowNames("Rfullfloat.bin"))
(cn<-GetJColNames("Rfullfloat.bin"))
# Get the names of rows and columns simultaneosuly as a list of two elements
(l<-GetJNames("Rfullfloat.bin"))
# Get several rows at once. The returned matrix has the rows in the
# same order as the passed list,
# and this list can contain even repeated values
(vm<-GetJManyRows("Rfullfloat.bin",c(1,4)))

# Of course, columns can be extrated equally
(vc<-GetJManyCols("Rfulldouble.bin",c(1,4)))
# and similar functions are provided for extracting by names:
(vm<-GetJManyRowsByNames("Rfulldouble.bin",c("A","D")))
(vc<-GetJManyColsByNames("Rfulldouble.bin",c("a","d")))
```

The package can manage and store sparse and symmetric matrices, too. 
```{r}
# Generation of a 6x8 sparse matrix
Rsp <- matrix(rep(0,48),nrow=6)
sparsity <- 0.1
nnz <- round(48*sparsity)
where <- floor(47*runif(nnz))
val <- runif(nnz)
for (i in 1:nnz)
{
 Rsp[floor(where[i]/8)+1,(where[i]%%8)+1] <- val[i]
}
rownames(Rsp) <- c("A","B","C","D","E","F")
colnames(Rsp) <- c("a","b","c","d","e","f","g","h")
# Let's see the matrix
Rsp
# Write the matrix as sparse with type float
JWriteBin(Rsp,"Rspafloat.bin",dtype="float",dmtype="sparse",
          comment="Sparse matrix of floats")
```

Notice that the condition of being a sparse matrix and the storage space 
used can be known with the matrix info.
```{r}
JMatInfo("Rspafloat.bin")
```

Be careful: trying to store as sparse a matrix which is not (it has not 
a majority of 0-entries) works, but produces
a matrix larger than the corresponding full matrix.

With respect to symmetric matrices, `JWriteBin` works the same way. 
Let us generate a $7 \times 7$ symmetric matrix.
```{r}
Rns <- matrix(runif(49),nrow=7)
Rsym <- 0.5*(Rns+t(Rns))
rownames(Rsym) <- c("A","B","C","D","E","F","G")
colnames(Rsym) <- c("a","b","c","d","e","f","g")
# Let's see the matrix
Rsym
# Write the matrix as symmetric with type float
JWriteBin(Rsym,"Rsymfloat.bin",dtype="float",dmtype="symmetric",
          comment="Symmetric matrix of floats")
# Get the information 
JMatInfo("Rsymfloat.bin")
```
Notice that if you store a R matrix which is NOT symmetric as a symmetric 
`jmatrix`, only the lower triangular part (including the
main diagonal) will be saved. The upper-triangular part will be lost.

The functions to read rows/colums stated before works equally independently 
of the matrix character (full, sparse or symmetric) so
you can play with them using the `Rspafloat.bin` and `Rsymfloat.bin` file 
to check they work.

If the jmatrix stored in a binary file has names associated to rows
or columns, you can filter it using them and generate another jmatrix file
with only the rows or columns you wish to keep. The function to do so is 'FilterJMatByName'.
```{r}
Rns <- matrix(runif(49),nrow=7)
rownames(Rns) <- c("A","B","C","D","E","F","G")
colnames(Rns) <- c("a","b","c","d","e","f","g")
# Let's see the matrix
Rns
# Write the matrix as full with type float
JWriteBin(Rns,"Rfullfloat.bin",dtype="float",dmtype="full",
          comment="Full matrix of floats")
# Extract the first two and the last two columns
FilterJMatByName("Rfullfloat.bin",c("a","b","f","g"),"Rfullfloat_fourcolumns.bin",namesat="cols")
# Let's load the matrix and let's see it
vm<-GetJManyRows("Rfullfloat_fourcolumns.bin",c(1,7))
vm
```


